Unifying intra- and inter-specific variation in tropical tree mortality

### 2018-08-09 ms/supplementary.Rnw
Tidy up SuppMatt, including past models

\title{Unifying intra- and inter-specific variation in tropical tree mortality}
\author{J.S. Camac, R. Condit, R.G. FitzJohn, L. McCalman, D. Steinberg, M. Westoby, S.J. Wright, D.S. Falster}
\correspondingauthor{Corresponding James S Camac.\\E-mail: james.camac@gmail.com}

@@

Here we provide some further details on methods and additional results, supplementing those presented in the main text. Section ``S1-Estimating True Growth'' describes a probabilistic methods used to estimate individual growth from field measurements, taking into account measurement error and the distribution of growth rate across the community. Section ``S2-Parameter distributions'' describes the distributions applied to species-level random effects and priors put over parameters in the fitted model. Section ``S3-Numerical optimisation'' outlines some numerical concerns in estimating logloss, while Section ``S4-Example Stan code'' gives an example of the code fitted to the data. Code and data reproducing the results in this paper are available at \href{https://github.com/traitecoevo/mortality_bci}{github.com/traitecoevo/mortality\_bci}.

@@

>>>>

\title{Unifying intra- and inter-specific variation in tropical tree mortality}
\author{J.S. Camac, R. Condit, R.G. FitzJohn, L. McCalman, D. Steinberg, M. Westoby, S.J. Wright, D.S. Falster}
\correspondingauthor{Corresponding James S Camac. E-mail: james.camac@gmail.com}

@@

Here we provide some further details on methods and additional results, supplementing those presented in the main text. 
The section ``Review of methods for estimating mortality'' 
``Estimating True Growth'' describes a probabilistic methods used to estimate individual growth from field measurements, taking into account measurement error and the distribution of growth rate across the community. 
The section ``Parameter distributions'' describes the distributions applied to species-level random effects and priors put over parameters in the fitted model. 
The section ``Numerical optimisation'' outlines some numerical concerns in estimating logloss, while 
the section ``Example Stan code'' gives an example of the code fitted to the data. Code and data reproducing the results in this paper are available at \href{https://github.com/traitecoevo/mortality_bci}{github.com/traitecoevo/mortality\_bci}.

\section*{Review of methods for estimating mortality}

While a variety of statistical methods have been applied to model tree mortality, these vary in two key ways.

First, is whether the time interval between census intervals is explicitly included in the model. Some methods model binary survival outcomes directly, while others seek to estimate the underlying mortality rate experienced by a population per unit time. Examples of the first group include logistic regression \cite[e.g.][]{monserud_modeling_1999, russo_interspecific_2008, vanMantgem:2009bi,dietze_tree_2011,}, generalized linear models with logit link \cite[e.g.][]{Visser_functional_2016}, and exponential or weibull models that don't explicitly include time \cite[e.g.][]{Kobe:1995tw, Wyckoff:2000un}. These methods are suitable for detecting significant correlations with predictors, or predicting survival over a fixed time interval. However, the second approach -- estimating hazard rates -- is generally preferable when formulating predictive survival models, as these models naturally incorporate census interval \cite{Kleinbaum:2012tn}. Models formulated around a hazard rate can therefore make predictions across different time intervals. Moreover, the hazard rate connects directly with the differential equations used to build process-based models of vegetation \cite[e.g.][]{Moorcroft:2001ws}. While a number of studies have directly estimated hazard rates, studies differ in the structure of the estimated hazard rate. Many studies estimate average hazard rates for populations of each species \cite[e.g.][]{Condit:2006, king:2006va, Kraft:2010kq, Wright:2010fl}. These studies are effective at quantifying differences between species but unable to explain variation among individuals within species. Others, seek to model variation in the hazard rate via a combination covariates \cite[e.g.][]{Ruger:2011cv} (see also main text).

The other way survival models may differ is in how the predicted survival curve responds to covariates. Both the hazard rate and survival models predict a relationship between covariates and the probability of death across a specified time period. In the main text, we argue that biologically, it makes sense for this probability to asymptote above zero. Yet, aside form the additive hazards model we present, the other methods that we have encountered will all drive probability of death to zero when growth rate is high (assuming growth rate is included as a covariate). In other words, the models behave like the centre panel in Fig 1b.

\subsection*{Logistic regression and Generalized linear models}

Studies using logistic regression, or Generalized linear models with a logit link , all seek to predict he probability of death over a specified time interval via a model of the form:
\begin{equation} \label{eq:logistic_model}
p_{i,t_1\rightarrow t_2} = \frac {1}{1+e^{ (\beta _{0} + \beta _{1}x_1 + \ldots )}},
\end{equation}
where $\beta_{0}, \beta _{1}, \dots$ are coefficients and $x_{1}, x_{2}, \dots$ are covariates. As increases in growth tend to decrease mortality risk, the coefficient for growth rate when used as a predictor will be positive. High growth rates will cause the exponential term to become large, and $p_{i,t_1\rightarrow t_2} \rightarrow 0$.

\subsection*{Proportional-hazards models}

Cox proportional-hazard models, or complementary-log-log link binomial regression \cite{Kleinbaum:2012tn}, both use a formulation for the hazard rate $\lambda_i(t)$ that takes the form 
\begin{equation} \label{eq:Cox_model}
\lambda_i(t)= e^{\beta _{0} + \beta _{1} x_1 + \ldots}.
\end{equation}
From eq 1 in the main text, the probability an individual dies is then
\begin{equation} \label{eq:mortality_model}
p_{i,t_1\rightarrow t_2} = 1 - \exp\left(- \int_{t_1}^{t_2} \lambda_i(t) {\rm d}t\right).
\end{equation}
As increases in growth rate tend to decrease mortality risk, the coefficient for growth rate when used as a covariate in eq. \ref{eq:Cox_model} will be negative. High growth rates will therefore cause $\lambda_i(t) \rightarrow 0$ and thus $p_{i,t_1\rightarrow t_2} \rightarrow 0$.

Another different but related approach has been to use a logistic function for $\lambda_i(t)$ \cite[e.g.][]{Ruger:2011cv}. This results in a similar outcome to the proportional-hazards models: high growth rates will cause $\lambda_i(t) \rightarrow 0$ and thus $p_{i,t_1\rightarrow t_2} \rightarrow 0$.

\subsection*{Other approaches}

Several papers model survival directly as an exponential function of form:
\begin{equation} \label{eq:exp_model}
p_{i,t_1\rightarrow t_2} = e^{\beta _{0} + \beta _{1} x_1 + \ldots},
\end{equation}
where $f$ is some function of covariates \cite[e.g.][]{Kobe:1995tw,Pacala:1996vm, Wyckoff:2000un, Hawkes:2000ib, Vieilledent:2010fv}. As increases in growth rate tend to decrease mortality risk, the coefficient for growth rate when used as a covariate in eq. \ref{eq:exp_model} will be negative. High growth rates then drive $p_{i,t_1\rightarrow t_2} \rightarrow 0$..

@@

\section*{Estimating species light demand}

A species' light demand was calculated using annual canopy census data collected during 1985\textendash1990 and 1990\textendash1995. The canopy census recorded, in all 5 by 5 m subplots across the 50 ha plot, the presence of leaf in six height intervals (0-2,2-5,5-10,10-20,20-30, \textgreater 30 m). The light demand for each species was estimated as the mean light index encountered by new saplings appearing in the census, as described in SI Appendix (see also Fig. \ref{figS2}).

For each subplot, we calculated the number of strata > 2 m containing vegetation; and then transformed this to a light index ranging from 0 (dense shade) to 1 (gap). As light may penetrate into a subplot from the edge of a subplot, we rescaled this index to account for values in the eight immediate neighbouring subplots. Specifically, we used a weighted sum approach whereby the central subplot is assigned a weight of 8 and the eight neighbouring subplots are assigned a weight of 1. This meant that the contribution of the central plot was equivalent to the combined effect of all eight neighbouring plots. These weighted values were then summed and rescaled between 0 and 1 by dividing by the maximum value estimated across all subplots.
